master thesis

### 2018-03-22 Paper - Latex/Paper - Skizze - v4.tex
Paper - Update Workflows; rearrange some Sections

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{OTL}
      For this project a large database for phylogenetic trees and also for a taxonomic tree is needed. 
        Since an ancestral state reconstruction algorithm is applied to the phylogenetic tree, and for 
        the assessment and other properties the taxonomy provides much more information. \\
      OTL gives a synthesis of phylogenetic trees (currently 819 trees) and a taxonomic tree\footnote{
          \hyperlink{https://tree.opentreeoflife.org/about/synthesis-release/v9.1}
          {https://tree.opentreeoflife.org/about/synthesis-release/v9.1}; Last checked: 22.03.2018.
        }. OTL also 
        includes the large phylogenetic database TreeBASE \cite{Hinchliff2015}. \\
      \todo{Das steht auf der Website nicht in dem Paper...} \\
      For phylogenetic data, there are at least five big data collections, namely:
      \begin{itemize}
        \item ITIS (Integrated Taxonomic Information System) \cite{ITIS}
        \item NCBI (National Center for Biotechnology Information) \cite{NCBI1988}
        \item WORMS (World Register of Marine Species) \cite{WoRMS2018}
        \item GBIF (Global Biodiversity Information Facility) \cite{GBIF}
        \item OTT (OpenTreeOfLife-Taxonomy) \cite{Hinchliff2015}
      \end{itemize}
      \todo{In die discussion?:} \\
      ITIS is only a small set of 100~\% confirmed and named species. GBIF is not composed with the help 
        of phylogeny, the same is valid for the NCBI taxonomy. The WORMS taxonomy is a way too small 
        dataset of mostly marine species. \\
      Here the taxonomy from OTL is used because it is including most of the known taxonomies and is 
        synthesised by preffering taxonomies that match with available phylogenetic data. Furthermore 
        the team from OTL preferre a maximum number of species \cite{Hinchliff2015}. This is resulting 
        in somekind of hybrid between taxonomy and phylogeny. \anmerkungstext{Wie genau ist das ein 
        Hybrid? Genauer beschreiben, was Du damit meinst... (Thilo)} \\

      A closer look is being made to some of the features of the synthesis tree. On the one hand the 
        distribution of the taxa and on the other the distribution of the nodes on the taxa. Since this 
        is not directly relevant to \colorbox{red}{our} study, there is a section in the appendix 
        \ref{sec:otl analysis}.

@@

      This database consists of entries of the form: species A (source) interacts with B (target). \\
      A number of interactions have been identified\footnote{\hyperlink{
        https://github.com/jhpoelen/eol-globi-data/blob/master/eol-globi-lib/src/main/java/org/eol/globi/domain/InteractType.java
        }{https://github.com/jhpoelen/eol-globi-data/.../InteractType.java}}
        ,including those that the species source or target has become a parasite or a free-living 
        species from the biological perspective. These are the following:
      \begin{itemize}
        \item free-living source: preysOn, eats, flowersVisitedBy, hasPathogen, pollinatedBy, 
          hasParasite, hostOf
        \item free-living target: preyedUponBy, parasiteOf, visitsFlowersOf, pathogenOf, hasHost
        \item parasite source: parasiteOf, pathogenOf
        \item parasite target: hasParasite, hasPathogen
      \end{itemize}
      Of these interactions, e.g. species A parasitize species B, the state of the species is 
        determined, here is species A parasitic and species B free-living. The case a parasite 
        conquers (parasitizes) another parasite yields conflicting states for the second species. 
        This is solved by preferring parasitic. \\
      For each species known IDs are stored in GloBI. This includes OTT (the taxonomy of OTL). All 
        species that have stored an OTT id and have a matching interaction are formed into two lists: 
        parasites and free-livings.

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Metadata analysis}
    In order to generate a realistic simulation, influencing parameters are investigated. \\
    There are two major types of parameters:
    \begin{enumerate}
      \item Biological parameters (A result of the evolutionary process.):
        \begin{itemize}
          \item transition probabilities
        \end{itemize}
      \item Distribution of the loss of information:
        \begin{itemize}
          \item Loss of topology ($\rightarrow$ mutlifurcations)
          \item Unknown information about states of some leaf nodes
        \end{itemize}
    \end{enumerate}
    The influence of these parameters are tested on our result using our simulation (Section 
      \ref{sec:simulation}).

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Transition probabilities}
      This subsection deals with the transition probabilities from free-living (hereinafter / as a 
        formula FL) to parasitic (hereinafter P) and vice versa: $\mathcal{P}(FL \rightarrow P)$, 
        $\mathcal{P}(P \rightarrow FL)$. \\
      Different parasite types have different transition probabilities. It is very difficult to make a 
        statement about these probabilities.
      In general, \colorbox{red}{we} assume that there are 40~\% parasites and 60~\% free-livings which 
        is based on the estimates by Windsor \cite{Windsor1998} and 
        $\mathcal{P}(FL \rightarrow P) > \mathcal{P}(P \rightarrow FL)$, because a reverse mutation is 
        usually less likely. \todo{This is discussed in section x of the discussion.} \\

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Simulation}\label{sec:simulation}
    There are various possibilities of ancestral state reconstruction. The simulation compares
      different algorithms with each other. \\
    First different implementations of the Fitch maximum parsimony are compared and then the best of 
      them is compared with the implementation of the sankoff algorithm of the castor package 
      \cite{Louca2017}. \\

@@

      %---------------------------------------------------------------------------------------------------
      %---------------------------------------------------------------------------------------------------
      \subsection{Fitch maximum parsimony}
        Fitch maximum parsimony is an algorithm for rooted, binary trees and describes an ancestral state 
          reconstruction for discrete states \cite{Fitch1971} by minimizing transitions between states. \\
        Note, the original Fitch algorithm has the sole purpose of minimizing the number of transitions 
          and not reconstructing the ancestral nodes. Felsenstein \cite{Felsenstein2003} describes a 
          simple extension for the reconstruction. Cunningham et al. \cite{Cunningham1998} have refined 
          these. \todotext{Wir haben mit ein paar kleinen änderungen optimiert... und schließlich auf 
          multifurcated angepasst...} \todotext{eigentlich ist Cunningham 'nur' eine kritische 
          Neubewertung. Sie beziehen ihren Algortihmus auf Swofford und Maddison...} \\
        To understand the differences to the multifurcated case, the algorithm for the binary case is 
          briefly explained and referred to the extension. \\

        Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
          a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
          Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
          has only one. Leaf nodes have their state as a set (eg \{FL\} or \{P\}, unknown leaf nodes the 
          union of all possible states (\{FL, P\}). \\
        The algorithm traverses three times trough the tree and fills these sets. \\
        In each step, two sets are considered and their intersect formed. There are two cases:
        \begin{enumerate}
          \item The intersection is not empty and corresponds to the new set.
          \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
        \end{enumerate}

        First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.
          \begin{wrapfigure}{r}{0.5\textwidth}
            \centering
            \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
            \caption{Fitch algorithm for binary trees. \\
              The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
              root node) consists of three sets, where the last set is the final one (bold). \\
              From the second internal node (seen from the root node) there are several possibilities to 
              create the second and third set.}
            \label{fig: binary Fitch}
          \end{wrapfigure} 
          Each internal node is formed from its child nodes, where at the beginning the only information
          lies. \\
        Second traverse from the root node to the leafs. Each internal node is formed from its father node 
          and its sibling node. \\
        Last traversion (direction does not matter): Build the final state for every node. It is formed 
          from the sets of previous traversals. \\
        (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
          of internal nodes, so it was just the first traversal.) \\
        The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
          more than two children may be present for the first traversal, but the incision or union may 
          also be formed over more than two sets. Also in the second traversing, there may be several 
          sibling nodes. However, there are several possibilities here that were all tested and compared 
          in the simulation. Some of these options are already available in the binary case:
        \begin{itemize}
          \item The father node has (except for the root node) two state sets, because he came through 
            the up-traversing previously. Are both sets used or only the first traversing?
          \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
            or directly in the whole with the father node?
        \end{itemize}
        The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
          both possibilities of the three sets. \\
        Cunningham uses only the first state set of the father node \cite{Cunningham1998}. \\
        From these two points four different versions of Fitch were formed:
        \begin{enumerate}
          \item Fitch 1: First state set of father node; intersection/union of siblings first.
          \item Fitch 2: First state set of father node; intersection/union of siblings together with father node.
          \item Fitch 3: Both state sets of father node; intersection/union of siblings first.
          \item Fitch 4: Both state sets of father node; intersection/union of siblings together with father node sets.
        \end{enumerate}
        \begin{figure}
          \centering
          \includegraphics[width=0.8\textwidth]{Figures/simulation_fitch_evaluation.png}
          \caption{Test of Fitch Versions.}
          \label{fig:Fitch versions}
        \end{figure}
        These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and a
          distribution of 60~\% FL to 40~\% P. Figure \ref{fig:Fitch versions} shows this over all unknown 
          node percentage. \\
        % At 90~\% unknown nodes and 90~\% of multifurcation of the internal nodes, version 1 was 89.26~\%, 
        %   version 2 was 89.26~\%, version 3 was 89.35~\%, and version 4 was 89.31~\% correct. Therefore, 
        %   only version 3 was used for all further simulations.
        At 95~\% unknown nodes and 95~\% of multifurcation of the internal nodes, version 1 was 88.37~\%, 
          version 2 was 88.37~\%, version 3 was 88.4~\%, and version 4 was 88.39~\% correct. Therefore, 
          only version 3 was used for all further simulations.

>>>>

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{OTL}
      For this project a large database for phylogenetic trees and also for a taxonomic tree is needed. 
        Since an ancestral state reconstruction algorithm is applied to the phylogenetic tree, and for 
        the assessment and other properties the taxonomy provides much more information. \\
      OTL gives a synthesis of phylogenetic trees (currently 819 trees) and a taxonomic tree\footnote{
          \hyperlink{https://tree.opentreeoflife.org/about/synthesis-release/v9.1}
          {https://tree.opentreeoflife.org/about/synthesis-release/v9.1}; Last checked: 22.03.2018.
        }. OTL also 
        includes the large phylogenetic database TreeBASE \cite{Hinchliff2015}. \\
      \todo{Das steht auf der Website nicht in dem Paper...} \\
      For phylogenetic data, there are at least five big data collections, namely:
      \begin{itemize}
        \item ITIS (Integrated Taxonomic Information System) \cite{ITIS}
        \item NCBI (National Center for Biotechnology Information) \cite{NCBI1988}
        \item WORMS (World Register of Marine Species) \cite{WoRMS2018}
        \item GBIF (Global Biodiversity Information Facility) \cite{GBIF}
        \item OTT (OpenTreeOfLife-Taxonomy) \cite{Hinchliff2015}
      \end{itemize}
      ITIS is only a small set of 100~\% confirmed and named species. GBIF is not composed with the help 
        of phylogeny, the same is valid for the NCBI taxonomy. The WORMS taxonomy is a way too small 
        dataset of mostly marine species. \\
      Here the taxonomy from OTL is used because it is including most of the known taxonomies and is 
        synthesised by preffering taxonomies that match with available phylogenetic data. Furthermore 
        the team from OTL preferre a maximum number of species \cite{Hinchliff2015}. This is resulting 
        in somekind of hybrid between taxonomy and phylogeny. \anmerkungstext{Wie genau ist das ein 
        Hybrid? Genauer beschreiben, was Du damit meinst... (Thilo)} \\

      A closer look is being made to some of the features of the synthesis tree. On the one hand the 
        distribution of the taxa and on the other the distribution of the nodes on the taxa. Since this 
        is not directly relevant for this study, there is a section in the appendix 
        \ref{sec:otl analysis}.

@@

      This database consists of entries of the form: species A (source) interacts with B (target). \\
      A number of interactions have been identified\footnote{\hyperlink{
        https://github.com/jhpoelen/eol-globi-data/blob/master/eol-globi-lib/src/main/java/org/eol/globi/domain/InteractType.java
        }{https://github.com/jhpoelen/eol-globi-data/.../InteractType.java}; Last checked: 22.03.2018.}
        , including those that the species source or target has become a parasite or a free-living 
        species from the biological perspective. These are the following:
      \begin{itemize}
        \item free-living source: preysOn, eats, flowersVisitedBy, hasPathogen, pollinatedBy, 
          hasParasite, hostOf
        \item free-living target: preyedUponBy, parasiteOf, visitsFlowersOf, pathogenOf, hasHost
        \item parasite source: parasiteOf, pathogenOf
        \item parasite target: hasParasite, hasPathogen
      \end{itemize}
      Of these interactions, e.g. species A parasitize species B, the state of the species is 
        determined, here is species A parasitic and species B free-living. The case a parasite 
        conquers (parasitizes) another parasite yields conflicting states for the second species. 
        This is solved by preferring parasitic. \\
      For each species known IDs are stored in GloBI. This includes OTT (the taxonomy of OTL). All 
        species that have stored an OTT id and have a matching interaction are formed into two lists: 
        parasites and free-livings.

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Metadata analysis}
    In order to generate a realistic simulation, influencing parameters are investigated. \\
    Since the transitions are minimized in an ancestral state reconstruction, this is an important 
      parameter to consider. On the other hand, the completeness of our input data are influencing 
      values. Therefore, two major types are distinguished:
    \begin{enumerate}
      \item Biological parameters (A result of the evolutionary process.):
        \begin{itemize}
          \item transition probabilities
        \end{itemize}
      \item Distribution of the loss of information:
        \begin{itemize}
          \item Loss of topology ($\rightarrow$ mutlifurcations)
          \item Unknown information about states of some leaf nodes
        \end{itemize}
    \end{enumerate}
    The influence of these parameters are tested on our result using our simulation (Section 
      \ref{sec:simulation}).

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Transition probabilities}
      As mentioned above, in an ancestral state reconstruction, transitions are minimized, with 
        transition probabilities playing a role. \\
      This subsection deals with these transition probabilities from free-living (hereinafter / as a 
        formula FL) to parasitic (hereinafter P) and vice versa: $\mathcal{P}(FL \rightarrow P)$, 
        $\mathcal{P}(P \rightarrow FL)$. \\
      Different parasite types have different transition probabilities. It is very difficult to make a 
        statement about these probabilities. It is generally assumed in this work that there are 40~\% 
        parasites and 60~\% free-livings which is based on the estimates by Windsor \cite{Windsor1998} 
        and $\mathcal{P}(FL \rightarrow P) > \mathcal{P}(P \rightarrow FL)$, because a reverse mutation 
        is usually less likely. \todo{This is discussed in section x of the discussion.} \\

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Fitch maximum parsimony}
      Fitch maximum parsimony is an algorithm for rooted, binary trees and describes an ancestral state 
        reconstruction for discrete states \cite{Fitch1971} by minimizing transitions between states. \\
      Note, the original Fitch algorithm has the sole purpose of minimizing the number of transitions 
        and not reconstructing the ancestral nodes. Felsenstein \cite{Felsenstein2003} describes a 
        simple extension for the reconstruction. Cunningham et al. \cite{Cunningham1998} have refined 
        these. \todotext{Wir haben mit ein paar kleinen änderungen optimiert... und schließlich auf 
        multifurcated angepasst...} \todotext{eigentlich ist Cunningham 'nur' eine kritische 
        Neubewertung. Sie beziehen ihren Algortihmus auf Swofford und Maddison...} \\
      To understand the differences to the multifurcated case, the algorithm for the binary case is 
        briefly explained and referred to the extension. \\
        \begin{wrapfigure}[19]{r}{0.5\textwidth}
          \centering
          \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
          \caption{Fitch algorithm for binary trees. \\
            The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
            root node) consists of three sets, where the last set is the final one (bold). \\
            From the second internal node (seen from the root node) there are several possibilities to 
            create the second and third set.}
          \label{fig: binary Fitch}
        \end{wrapfigure} 
      Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
        a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
        Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
        has only one. Leaf nodes have their state as a set (eg \{FL\} or \{P\}, unknown leaf nodes the 
        union of all possible states (\{FL, P\}). \\

      The algorithm traverses three times trough the tree and fills these sets. \\
      In each step, two sets are considered and their intersect formed. There are two cases:
      \begin{enumerate}
        \item The intersection is not empty and corresponds to the new set.
        \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
      \end{enumerate}
      First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.
        Each internal node is formed from its child nodes, where at the beginning the only information
        lies. \\
      Second traverse from the root node to the leafs. Each internal node is formed from its father node 
        and its sibling node. \\
      Last traversion (direction does not matter): Build the final state for every node. It is formed 
        from the sets of previous traversals. \\
      (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
        of internal nodes, so it was just the first traversal.) \\
      The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
        more than two children may be present for the first traversal, but the incision or union may 
        also be formed over more than two sets. Also in the second traversing, there may be several 
        sibling nodes. However, there are several possibilities here that were all tested and compared 
        in the simulation. Some of these options are already available in the binary case:
      \begin{itemize}
        \item The father node has (except for the root node) two state sets, because he came through 
          the up-traversing previously. Are both sets used or only the first traversing?
        \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
          or directly in the whole with the father node?
      \end{itemize}
      The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
        both possibilities of the three sets. \\
      Cunningham uses only the first state set of the father node \cite{Cunningham1998}. \\
      From these two points four different versions of Fitch were formed:
      \begin{enumerate}
        \item Fitch 1: First state set of father node; intersection/union of siblings first.
        \item Fitch 2: First state set of father node; intersection/union of siblings together with father node.
        \item Fitch 3: Both state sets of father node; intersection/union of siblings first.
        \item Fitch 4: Both state sets of father node; intersection/union of siblings together with father node sets.
      \end{enumerate}
      \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{Figures/simulation_fitch_evaluation.png}
        \caption{Test of Fitch Versions.}
        \label{fig:Fitch versions}
      \end{figure}
      These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and a
        distribution of 60~\% FL to 40~\% P. Figure \ref{fig:Fitch versions} shows this over all unknown 
        node percentage. \\
      % At 90~\% unknown nodes and 90~\% of multifurcation of the internal nodes, version 1 was 89.26~\%, 
      %   version 2 was 89.26~\%, version 3 was 89.35~\%, and version 4 was 89.31~\% correct. Therefore, 
      %   only version 3 was used for all further simulations.
      At 95~\% unknown nodes and 95~\% of multifurcation of the internal nodes, version 1 was 88.37~\%, 
        version 2 was 88.37~\%, version 3 was 88.4~\%, and version 4 was 88.39~\% correct. Therefore, 
        only version 3 was used for all further simulations.

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Simulation}\label{sec:simulation}
    The simulation compares these different of ancestral state reconstruction algorithms with each 
      other. \\
    First different implementations of the Fitch maximum parsimony are compared and then the best of 
      them is compared with the implementation of the Sankoff algorithm of the \textit{Castor} package 
      \cite{Louca2017}. \\

@@
